Discussion Draft

### 2020-04-03 doc/Discussion.rmd
some work on rec landscape

Directional selection could be the result in spermatogenesis if the fitness landscape for 'chromosome structure' is altered via genetic architecture changes or GxE effects. (in oogenesis the leaky SAC has weak selection coefficient)

@@

The spindle based model proposed that sexual dimorphism in the recombination landscape is driven by difference strength in the spindle assembly checkpoint (Lane Kauppi). The relaxed selection imposed by an ineffective SAC results in a uniform distribution of positions (while maintaining positive interference).

Just as the difference in SAC (checkpoint) strength can lead to disinct evolutionary trajectories of the genome wide recombination rate for oocytes and spermatocytes, the other side of the same coin is that there is increased variance between cells for crossover number. (weaker / leaky checkpoint )

The telomeric placement of crossovers in the male recombiation landscape, minimizes the amount of sister cohesion connecting homologs. This conserved feature of the male recombination landscape, might act to synchornize the seperation homologs after alignment at metaphase, (and in the degargation of the sister cohesion connecting the homologs. Another conserved feature of gametogenesis is that the timing of spermatocyte metapphase is much shorter of time compared to oocytes.

>>>>

<Directional selection could be the result in spermatogenesis if the fitness landscape for 'chromosome structure' is altered via genetic architecture changes or GxE effects. (in oogenesis the leaky SAC has weak selection coefficient) ... bret says remove>

@@

**sister cohesin, chromosome structure altered by the number and placement of crossovers** synchornize timing of the decay of sister cohesion connecting homologs at metaphase I. conserved gametogensis differences, spermatogenesis faster than oogenesis -- The sex difference in requirements for sister cohesion could hinge on the timing differences between spermatogenesis and oogenesis.

The connection of the **The spindle based model** centers on sex differences on the requirements for chromosome cohesion in late meiosis I.  In oocytes the slower cell cycle and prolonged arrest may require chromosome structures with greater stability on the MI spindle.

The number and placement of crossovers alters the lay-out / distribution of sister cohesion and the 'chromosome stucture' once kinetochores attached to MT-spindle and tension applied.  For single acrocentric chromsoomes the sister cohesion (area of the chromosome) distal to the crossover connects the homologs. (known as cruciform tetrad is some systems). (this is the part of the tension force read by the SAC).

This area is reduced when crossover position is shifted towards the telomere (ends of the chromsome). The transition of metaphase to anaphase proceeds by the (irreversable) proteolytic decay of the sister cohesion on these tetrad arms. The male telomere bias locatlization of crossovers could be adaptive in order to synchornize (or minimize) the timing of homologs sepearting at metaphase I.

 Another conserved feature of gametogenesis is that the timing of spermatocyte metapphase is much shorter of time compared to oocytes.
